---
title: Programming in R
date: 2026-01-06
date-modified: 2026-01-06
---

<!-- Some math definitions -->
\renewcommand{\Pr}[1]{\text{Pr}\left(#1\right)}
\newcommand{\Prcond}[2]{\Pr{\left.\vphantom{#2}#1\right|#2}}

The R programming language [@R] is the de facto language for social network analysis[^notfornetsci]. Furthermore, R is home to the most comprehensive collection of packages implementing the methods we will cover here. Let's start with the fundamentals.

[^notfornetsci]: This is less true for network science in general, where Python is more prevalent.

## Getting help

Unlike other languages, R's documentation is highly reliable. The Comprehensive R Archive Network [CRAN] is the official repository of R packages. All packages posted on CRAN must pass a series of tests to ensure the quality of the code, including the documentation. 

To get help on a function, we can use the `help()` function. For example, if we wanted to get help on the `mean()` function, we would do:

```{r}
#| label: help
#| eval: false
help("mean")
```

## Naming conventions

R has a set of naming conventions that we should follow to avoid confusion. The most important ones are:

- Use lowercase letters (optional)
- Use underscores to separate words (optional)
- Do not start with a number
- Do not use special characters 
- Do not use reserved words

::: {.callout title="Question"}
Of the following list, which are valid names and which are valid but to be avoided?

```r
_my.var
my.var
my_var
myVar
myVar1
1myVar
my var
my-var
```
:::

## Assignment

In R, we have two (four) ways of assigning values to objects: the `<-` and `=` binary operators[^binop]. Although both are equivalent, the former is the preferred way of assigning values to objects since the latter can be confused with function arguments.

[^binop]: In mathematics and computer science, a binary operator is a function that takes two arguments. In R, binary operators are implemented as `variable 1` `[operator]` `variable 2`. For example, `1 + 2` is a binary operation.

```{r}
#| label: assignment
#| collapse: true
x <- 1
x = 1
```

::: {.callout title="Question"}
What is the difference between the following two assignments? Use the help function to find out.

```r
x <- 1
x <<- 1
```
:::

::: {.callout title="Question"}
What are other ways in which you can assign values to objects?
:::

## Using functions and piping

In R, we use functions to perform operations on objects. Functions are implemented as `function_name` `(` `argument_1` `,` `argument_2` `,` `...` `)`. For example, the `mean()` function takes a vector of numbers and returns the mean of the values:

```{r}
#| label: using-functions
#| collapse: true
x <- c(1, 2, 3) # The c() function creates a vector
mean(x)
```

Furthermore, we can use the pipe operator (`|>`) to improve readability. The pipe operator takes the output of the left-hand side expression and passes it as the first argument of the right-hand side expression. Our previous example could be rewritten as:

```{r}
#| label: using-functions-pipe
#| collapse: true
c(1, 2, 3) |> mean()
```


## Data structures

Atomic types are the minimal building blocks of R. They are logical, integer, double, character, complex, raw:

```{r}
#| label: data-structures
x_logical   <- TRUE
x_integer   <- 1L
x_double    <- 1.0
x_character <- "a"
x_complex   <- 1i
x_raw       <- charToRaw("a")
```

Unlike other languages, we do not need to declare the data type before creating the object; R will infer it from the value.

::: {.callout .callout-tip title="Pro-tip"}
Adding the `L` suffix to the value is good practice when dealing with integers. Some R packages like `data.table` [@datatable] have internal checks that will throw an error if you are not explicit about the data type.
:::

The next type is the vector. A vector is a collection of elements of the same type. The most common way to create a vector is with the `c()` function:

```{r}
#| label: data-structures-vectors
x_integer <- c(1, 2, 3)
x_double  <- c(1.0, 2.0, 3.0)
x_logical <- c(TRUE, FALSE, TRUE)
## etc.
```

R will coerce the data types to the most general type. For example, if we mix integers and doubles, R will coerce the integers into doubles. The coercion order is logical < integer < double < character

::: {.callout title="Question"}
Why is the coercion order logical < integer < double < character?
:::

The next data structure is the list. A list is a collection of elements of any type. We can create a list with the `list()` function:

```{r}
#| label: data-structures-lists
x_list       <- list(1, 2.0, TRUE, "a")
x_list_named <- list(a = 1, b = 2.0, c = TRUE, d = "a")
```

To access elements in a list, we have two options: by position or by name, the latter only if the elements are named:

```{r}
#| label: data-structures-lists-access
#| collapse: true
x_list[[1]]
x_list_named[["a"]]
x_list_named$a
```

After lists, we have matrices. A matrix is a collection of elements of the same type arranged in a two-dimensional grid. We can create a matrix with the `matrix()` function:

```{r}
#| label: data-structures-matrices
#| collapse: true
x_matrix <- matrix(1:9, nrow = 3, ncol = 3)
x_matrix

## We can access elements in a matrix by row column, or position:
x_matrix[1, 2]
x_matrix[cbind(1, 2)]
x_matrix[4]
```

::: {.callout title="Matrix is a vector"}
Matrices in R are vectors with dimensions. In base R, matrices are stored in column-major order. This means that the elements are stored column by column. This is important to know when we are accessing elements in a matrix
:::

The two last data structures are arrays and data frames. An array is a collection of elements of the same type arranged in a multi-dimensional grid. We can create an array with the `array()` function:

```{r}
#| label: data-structures-arrays
#| collapse: true
x_array <- array(1:27, dim = c(3, 3, 3))

## We can access elements in an array by row, column, and dimension, or
## position:
x_array[1, 2, 3]
x_array[cbind(1, 2, 3)]
x_array[22]
```

Data frames are the most common data structure in R. In principle, these objects are lists of vectors of the same length, each vector representing a column. Columns (lists) in data frames can be of different types, but elements in each column must be of the same type. We can create a data frame with the `data.frame()` function:

```{r}
#| label: data-structures-data-frames
#| collapse: true
x_data_frame <- data.frame(
  a = 1:3,
  b = c("a", "b", "c"),
  c = c(TRUE, FALSE, TRUE)
)

## We can access elements in a data frame by row, column, or position:
x_data_frame[1, 2]
x_data_frame[cbind(1, 2)]
x_data_frame$b[1]    # Like a list
x_data_frame[[2]][1] # Like a list too
```

## Functions

Functions are the most important building blocks of R. A function is a set of instructions that takes one or more inputs and returns one or more outputs. We can create a function with the `function()` function:

```{r}
#| label: functions
#| collapse: true
## This function has two arguments (y is optional)
f <- function(x, y = 1) {
  x + 1
}

f(1)
```

Starting with R 4, we can use the lambda syntax to create functions:

```{r}
#| label: functions-lambda
#| collapse: true
f <- \(x, y) x + 1

f(1)
```

## Control flow

Control flow statements allow us to control the execution of the code. The most common control flow statements are `if`, `for`, `while`, and `repeat`. We can create a control flow statement with the `if()`, `for()`, `while()`, and `repeat()` functions:

```{r}
#| label: control-flow
#| collapse: true
## if
if (TRUE) {
  "a"
} else {
  "b"
}

## for
for (i in 1:3) {
  cat("This is the number ", i, "\n")
}

## while
i <- 1
while (i <= 3) {
  cat("This is the number ", i, "\n")
  i <- i + 1
}

## repeat
i <- 1
repeat {
  cat("This is the number ", i, "\n")
  i <- i + 1
  if (i > 3) {
    break
  }
}
```

## R packages

R is so powerful because of its extensions. R extensions (different from other programming languages) are called packages. Packages are collections of functions, data, and documentation that provide additional functionality to R. Although anyone can create and distribute R packages to other users, the Comprehensive R Archive Network [CRAN] is the official repository of R packages. All packages posted on CRAN are thoroughly tested, so generally, their quality is high.

To install R packages, we use the `install.packages()` function; to load them, we use the `library()` function. For example, the following code chunk installs the `ergm` package and loads it:

```{r}
#| label: packages
#| collapse: true
#| eval: false
install.packages("ergm")
library(ergm)
```

# Using AI

